Estimating duration in the presence of misclassification

### 2024-03-09 introduction.tex
edit new structure

In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak.
Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
In all cases, surveillance is required to inform public health policy.
In this thesis, I consider methods to improve surveillance, using the SARS-SARS-CoV-2 epidemic in England as a case study.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is at least as important as how it is changing~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to estimating the parameters of this process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

@@

@@

\section{Transmission model}

The first component required to estimate incidence and transmission is a \emph{transmission model}, a mathematical description of how incidence arises.
A spectrum of models exist, categorised by the strenght of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

The models with the weakest assumptions are statistical models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process.
The most assumption is that incidence follows a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), and the intensity of the process is the target of incidence.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
\emph{Compartmental models} are the most common type of mechanistic model.
Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

@@

Typically, the core of an observation model is a \emph{convolution equation} of an approximation of it.
This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
\end{align}
where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.

@@

Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.

For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.

@@

\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
It combines a statistical transmission model with

@@

The application I will consider in this thesis is the SARS-CoV-2 epidemic in England.

\subsection{Data sources}

\subsubsection{For estimating incidence and transmission}

@@

\subsubsection{Cohort studies}

In this thesis, I make use of two cohort studies with differing designs.
These designs have complementary strengths, and I combine information from the studies to make use of these.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

@@

The second study is the CIS, which I previously mentioned.
In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

The \emph{backcalculation} framework~\autocite[e.g.][]{brookmeyerMeasuring,brookmeyerMethod,brookmeyerBackcalculation}, developed in the context of HIV/AIDS
\todo{this is a bad place for an intro to backcalculation: where does it go in this structure?}

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

The estimates I present in \cref{E-backcalc} are methodologically similar.
However, I fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study.
\Textcite{abbottCISincidence} include the short period of low but increasing test sensitivity at the start of an infection (see \cref{E-biology-data:sec:natural-history}).

\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
Their methodology was similar to \textcite{abbottCISincidence} (see \cref{intro:sec:backcalc}), including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.

Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.

>>>>

In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak.
Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
In all cases, surveillance is required to inform public health policy.
In this thesis, I consider methods to improve surveillance, using the SARS-CoV-2 epidemic in England as a case study.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required.
For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to estimating the parameters of this process, the infection process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

@@

\todo[inline]{Not sure this section (data sources) needs to exist, particularly in this location}
\todo[inline]{If this section kept, probably want to cite some things on generic data for pandemic surveillance}

@@

\section{Infection process model}

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of how incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

The models with the weakest assumptions are statistical models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of incidence.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
\end{align}
where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
\emph{Compartmental models} are the most common type of mechanistic model.
Compartmental models split the population into compartments, each representing a different state of the disease (described in full in \cref{E-SEIR:sec:mechanistic-models}).
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular the age structure of the population is important for respiratory diseases.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

\todo[inline]{Needs some generic citations about observation models}

@@

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
\end{align}
where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.

@@

\todo[inline]{Expand slightly on convolution equations and why solving them is challenging, including some citations}
Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.

@@

\todo[inline]{
  Decide whether to keep this section.
  I think backcalculation probably does need to be called out explicitly, but it feels a little out-of-place.
  IF kept, needs expanding a bit, flow a bit better, and citations added.
}

\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
It combines a statistical transmission model with an observation model for prevalence data.

Here, $\zeta = 1$ and $f(t)$ represents the probability of having the disease $t$ days after infection.
In this way, $f$ represents a duration.

\todo[inline]{Following paragraph seems a bit specific but is needed to motivate the following section}
For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.

@@

\subsection{Data for estimating incidence and transmission} \label{intro:sec:data-for-estimating}

@@

\subsection{Cohort studies}

In this thesis, I make use of two cohort studies with differing designs to estimate duration.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

@@

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.

Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
% These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
% The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
The duration of positivity in these models is limited by the model structure.
The models are fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
% These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
